Semantic integration of physiology phenotypes: an application to the cellular phenotype ontology

### 2012-01-19 trunk/paper/drinking-sr.tex

\documentclass{bioinfo}
\copyrightyear{2005}
\pubyear{2005}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{pifont}
%\newcommand{\tick}{\ding{52}}
\newcommand{\tick}{\checkmark}
\newcommand{\tickNo}{\hspace{1pt}\ding{55}}

\begin{document}
\firstpage{1}

\title[Cell Phenotype Ontology]{Semantic integration of physiology
  phenotypes: an application to the cellular phenotype ontology}

\author[Hoehndorf \textit{et~al}]{Robert Hoehndorf$^{1}$\footnote{to
    whom correspondence should be addressed}, Midori
  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and
  Georgios V. Gkoutos$^{1}$}

\address{$^{1}$Department of Genetics, University of Cambridge,
  Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
  $^{2}$Department of Biochemistry; University of Cambridge, 80 Tennis
  Court Road, Cambridge CB2 1GA, UK\\
  $^{3}$Institute for Medical Informatics, Statistics and
  Epidemiology, University of Leipzig, Haertelstrasse 16-18, 04107
  Leipzig, Germany\\
  $^{4}$European Bioinformatics Institute, Wellcome Trust Genome
  Campus, Hinxton, Cambridge, Cambridge CB10 1SD, UK}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

@@

\begin{abstract}
\section{Motivation:}
The systematic observation of phenotypes has become a crucial tool of
functional genomics, and several large international projects are
currently underway to identify and characterize the phenotypes that
are associated with genotypes in several species. To integrate
phenotype descriptions within and across species, phenotype ontologies
have been developed. Applying ontologies to unify phenotype
descriptions in the domain of physiology has been a particular
challenge due to the high complexity of the underlying domain.

\section{Results:}
Here, we present the outline of a theory for an ontology of
physiology-related phenotypes. We provide a formal description of
process attributes and relate them to the attributes of their parts
and participants. We apply our theory to create the Cellular Phenotype
Ontology (CPO). The CPO is an ontology of morphological and
physiological phenotypic abnormalities of cells, cell components and
cellular processes. Its prime application is the unification of
cellular phenotype descriptions across species by providing terms and
uniform definition patterns. The CPO can further be used for the
annotation of observed abnormalities in domains, such as systems
microscopy, in which cellular abnormalities are observed and for which
no phenotype ontology has been created.

\section{Availability and implementation:} 
The CPO and the source code we generated to create the CPO are freely
available on \url{http://cell-phenotype.googlecode.com}.

\section{Contact:} \href{rh497@cam.ac.uk}{rh497@cam.ac.uk}
\end{abstract}

\section{Introduction}
Phenotype studies on all scales and levels of granularity are now an
invaluable tool for functional genomics research. Phenotypes of
targeted mutations in animal models are now systematically recorded to
reveal the role of individual genes within a biological system. These
phenotype studies now play a key role in translational research and
are being used to reveal candidate genes for orphan diseases and to
identify chemicals that may have effects on these diseases
\cite{Schofield2011}.

@@

\begin{methods}
\section{System and methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable effort and
resources in the development and establishment of ontologies that are
becoming increasingly successful as information management and
integration tools in many disparate scientific fields allowing
interoperability and semantic information processing between diverse
biomedical resources and domains.

\end{methods}

@@

\begin{table*}
  \centering
  \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
    & Increased flow rate & Normal flow rate & Decreased
    flow rate \\
    \hline
    Increased frequency &increased total flow rate &increased total
    flow rate &?\\
    Normal frequency &increased total flow rate &normal total flow
    rate &decreased total flow rate\\
    Decreased frequency &?&decreased total flow rate &decreased total
    flow rate\\
    \hline
  \end{tabular}
  \caption{\label{tbl:flow}Interdependency for the attribute {\em
      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
      rate} is an attribute of {\em Regulation of cytoplasmic
      streaming} processes, while {\em Cytoplasmic flow rate} is an
    attribute of individual {\em cytoplasmic streaming}
    processes. Depending both on whether the cytoplasmic flow rate in
    individual {\em cytoplasmic streaming} processes is increased or
    decreased and whether the frequency of occurrence of {\em
      cytoplasmic streaming} is increased or decreased, the total
    cytoplasmic flow rate can be increased or decreased.}
\end{table*}

@@

\section{Discussion}
\subsection{Applications of the CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO-BP's {\em Cellular process} as the entity; a further
11\% reference GO-CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration} and {\em
  decreased occurrence}. FYPO will thus fit neatly under the CPO
umbrella, and stands to benefit from the automated synchronization
between CPO and GO, as well as the integration of cellular phenotypes
across species that the CPO can provide. {\em Schizosaccharomyces
  pombe} annotations to FYPO terms will provide a rich body of highly
specific, well-supported data to be integrated with data from other
species.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A further domain that will greatly benefit from the CPO is {\em
  systems microscopy}, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
\cite{Lock2010}.  If we consider some of the studies, which have been
published in this field in the last few years \cite{Neumann2010,
  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
three studies illustrated in Table \ref{tbl:studies}, live-cell
imaging assays and RNAi knockdown were used to generate phenotypic
profiles that quantify the cellular response to a given siRNA thus
allowing identification of hundreds of genes involved in diverse
biological functions including cell division, migration and survival.
In each study, several phenotypes were detected and described by the
authors without the use of ontologies (see Table \ref{tbl:studies}),
making the integration between datasets extremely difficult. For
example, it is evident that cell division phenotypes were observed in
all three datasets (e.g. {\em Mitotic delay/arrest} and {\em Prolonged
  mitotic exit} or {\em Methaphase delay} and {\em Methaphase cells})
but the overlap between such phenotypes is unclear.

Data integration is also complicated by the lack of standardization at
the level of data production and processing; all these issues are
currently being address by the different groups involved in the
Systems Microscopy Network of Excellence
(http://www.systemsmicroscopy.eu/) and the first step towards data
integration can be achieved by further developing CPO.

@@

\subsection{Future research}
We implemented the CPO using a pattern-based approach to formulating
phenotypes involving processes. The patterns we identify are based on
pre-existing ontologies, in particular the PATO ontology and the
classification of cellular processes as well as cellular components in
the GO. The result of our method is a large ontology in which classes
for phenotypes are {\em pre-composed}: they are named and defined
within an OWL ontology. However, the large size of the resulting
ontology may impair its utility for data annotation and integration,
and software tools may not always support such very large
ontologies. The alternative to pre-composing all possible phenotype
classes using the patterns we describe is to dynamically generate
appropriately defined classes at the time at which they are being
used. To achieve this goal, software must be developed to support
ontology users in applying these patterns and generate the appropriate
class description when required.

@@

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{onto.pdf}
  \caption{Overview over basic top-level categories. Physical objects
    are entities that are wholly present at time points. Qualities are
    attributes of entities. Functions are capabilities that arise from
    physical objects with their qualities, and processes are
    temporally extended entities that may be realizations of a
    function.\label{fig:onto}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{processpatterns.pdf}

@@

\begin{figure*}
  \centering
  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

\section{Acknowledgement}
Funding for RH is provided by the European Commission's 7th Framework
Programme, RICORDO project, grant number 248502. MAH is supported by
Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
Medical Informatics, Statistics and Epidemiology, University of
Leipzig.  Funding for GR is provided by the European Union's Seventh
Framework Programme (FP7/2007-2013) under grant agreement number
258068, EU-FP7-Systems Microscopy NoE.

\bibliographystyle{plain}
\bibliography{lc}

>>>>

\documentclass[11pt]{article}
%\usepackage{times}
\usepackage{setspace}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
%\copyrightyear{2005}
%\pubyear{2005}
\usepackage{graphicx}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{pifont}
% Load packages
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage[utf8]{inputenc} %unicode support
\urlstyle{rm}
\usepackage[sort&compress]{natbib}
\renewcommand{\cite}{\citep}
\usepackage{amsmath}
\usepackage{amssymb}

\begin{document}

\title{Semantic integration of physiology
  phenotypes: an application to the cellular phenotype ontology}

\author{Robert Hoehndorf$^{1,*}$, Midori
  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and\\
  Georgios V. Gkoutos$^{1}$}

\date{$^{1}$Department of Genetics, University of Cambridge,
  Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
  $^{2}$Department of Biochemistry; University of Cambridge, 80 Tennis
  Court Road, Cambridge CB2 1GA, UK\\
  $^{3}$Institute for Medical Informatics, Statistics and
  Epidemiology, University of Leipzig, Haertelstrasse 16-18, 04107
  Leipzig, Germany\\
  $^{4}$European Bioinformatics Institute, Wellcome Trust Genome
  Campus, Hinxton, Cambridge, Cambridge CB10 1SD, UK}

% \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

% \editor{Associate Editor: XXXXXXX}

@@

\begin{abstract}
The systematic observation of phenotypes has become a crucial tool of
functional genomics, and several large international projects are
currently underway to identify and characterize the phenotypes that
are associated with genotypes in several species. To integrate
phenotype descriptions within and across species, phenotype ontologies
have been developed. Applying ontologies to unify phenotype
descriptions in the domain of physiology has been a particular
challenge due to the high complexity of the underlying domain.
Here, we present the outline of a theory for an ontology of
physiology-related phenotypes. We provide a formal description of
process attributes and relate them to the attributes of their parts
and participants. We apply our theory to create the Cellular Phenotype
Ontology (CPO). The CPO is an ontology of morphological and
physiological phenotypic abnormalities of cells, cell components and
cellular processes. Its prime application is the unification of
cellular phenotype descriptions across species by providing terms and
uniform definition patterns. The CPO can further be used for the
annotation of observed abnormalities in domains, such as systems
microscopy, in which cellular abnormalities are observed and for which
no phenotype ontology has been created.  
The CPO and the source code we generated to create the CPO are freely
available on \url{http://cell-phenotype.googlecode.com}.
\end{abstract}

%\section{Introduction}
Phenotype studies on all scales and levels of granularity are now an
invaluable tool for functional genomics research. Phenotypes of
targeted mutations in animal models are now systematically recorded to
reveal the role of individual genes within a biological system. These
phenotype studies now play a key role in translational research and
are being used to reveal candidate genes for orphan diseases and to
identify chemicals that may have effects on these diseases
\cite{Schofield2011}.

@@

@@

@@

\section{Discussion}
%\subsection{Applications of the CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO-BP's {\em Cellular process} as the entity; a further
11\% reference GO-CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration} and {\em
  decreased occurrence}. FYPO will thus fit neatly under the CPO
umbrella, and stands to benefit from the automated synchronization
between CPO and GO, as well as the integration of cellular phenotypes
across species that the CPO can provide. {\em Schizosaccharomyces
  pombe} annotations to FYPO terms will provide a rich body of highly
specific, well-supported data to be integrated with data from other
species.

A further domain that will greatly benefit from the CPO is {\em
  systems microscopy}, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
\cite{Lock2010}.  If we consider some of the studies, which have been
published in this field in the last few years \cite{Neumann2010,
  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
three studies, live-cell imaging assays and RNAi knockdown were used
to generate phenotypic profiles that quantify the cellular response to
a given siRNA thus allowing identification of hundreds of genes
involved in diverse biological functions including cell division,
migration and survival.  In each study, several phenotypes were
detected and described by the authors without the use of ontologies or
controlled vocabularies, making the integration between datasets
extremely difficult. For example, it is evident that cell division
phenotypes were observed in all three datasets and describes using
terms such as {\em Mitotic delay/arrest}, {\em Prolonged mitotic
  exit}, {\em Methaphase delay} and {\em Methaphase cells}). the
overlap between such phenotypes is unclear.

Data integration is also complicated by the lack of standardization at
the level of data production and processing; all these issues are
currently being address by the different groups involved in the
Systems Microscopy Network of Excellence
(\url{http://www.systemsmicroscopy.eu/}) and the first step towards
data integration can be achieved by further developing CPO.

@@

%\subsection{Future research}
We implemented the CPO using a pattern-based approach to formulating
phenotypes involving processes. The patterns we identify are based on
pre-existing ontologies, in particular the PATO ontology and the
classification of cellular processes as well as cellular components in
the GO. The result of our method is a large ontology in which classes
for phenotypes are {\em pre-composed}: they are named and defined
within an OWL ontology. However, the large size of the resulting
ontology may impair its utility for data annotation and integration,
and software tools may not always support such very large
ontologies. The alternative to pre-composing all possible phenotype
classes using the patterns we describe is to dynamically generate
appropriately defined classes at the time at which they are being
used. To achieve this goal, software must be developed to support
ontology users in applying these patterns and generate the appropriate
class description when required.

@@

\section{Methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable effort and
resources in the development and establishment of ontologies that are
becoming increasingly successful as information management and
integration tools in many disparate scientific fields allowing
interoperability and semantic information processing between diverse
biomedical resources and domains.

\bibliographystyle{unsrt}
\bibliography{lc}

\clearpage
\section*{Acknowledgement}
Funding for RH is provided by the European Commission's 7th Framework
Programme, RICORDO project, grant number 248502. MAH is supported by
Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
Medical Informatics, Statistics and Epidemiology, University of
Leipzig.  Funding for GR is provided by the European Union's Seventh
Framework Programme (FP7/2007-2013) under grant agreement number
258068, EU-FP7-Systems Microscopy NoE.

\section*{Author contributions}
...

\section*{Additional information}
\paragraph*{Competing financial interests:} The author(s) declare no
competing financial interests.

\clearpage
\section*{Figure legends}
\begin{figure}[h]
  \centering
%  \includegraphics[width=1\textwidth]{onto.pdf}
  \caption{Overview over basic top-level categories. Physical objects
    are entities that are wholly present at time points. Qualities are
    attributes of entities. Functions are capabilities that arise from
    physical objects with their qualities, and processes are
    temporally extended entities that may be realizations of a
    function.\label{fig:onto}}
\end{figure}

\begin{figure}[h]
  \centering
 % \includegraphics[width=1\textwidth]{processpatterns.pdf}

@@

\begin{figure*}[h]
  \centering
%  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

\clearpage
\section*{Tables}
\begin{table*}[h]
  \centering
  \begin{tabular}{p{2.5cm}|p{3cm}|p{3cm}|p{3cm}}
    & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate &
    Decreased cytoplasmic flow rate \\
    \hline
    Increased frequency &increased total flow rate &increased total
    flow rate &?\\
    Normal frequency &increased total flow rate &normal total flow
    rate &decreased total flow rate\\
    Decreased frequency &?&decreased total flow rate &decreased total
    flow rate\\
    \hline
  \end{tabular}
  \caption{\label{tbl:flow}Interdependency for the attribute {\em
      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
      rate} is an attribute of {\em Regulation of cytoplasmic
      streaming} processes, while {\em Cytoplasmic flow rate} is an
    attribute of individual {\em cytoplasmic streaming}
    processes. Depending both on whether the cytoplasmic flow rate in
    individual {\em cytoplasmic streaming} processes is increased or
    decreased and whether the frequency of occurrence of {\em
      cytoplasmic streaming} is increased or decreased, the total
    cytoplasmic flow rate can be increased or decreased.}
\end{table*}
